---
title: "MCristinaFernandez.module05RProject"
author: "M. Cristina Fernandez"
date: "`r Sys.Date()`"
output: html_document
---

# The Situation

Your CEO has decided that the company needs a full-time data scientist, and possibly a team of them in the future. 
She thinks she needs someone who can help drive data science within then entire organization and could potentially lead a team in the future. 
She understands that data scientist salaries vary widely across the world and is unsure what to pay them. 
To complicate matters, salaries are going up due to the great recession and the market is highly competitive. 
Your CEO has asked you to prepare an analysis on data science salaries and provide them with a range to be competitive and get top talent. 
The position can work offshore, but the CEO would like to know what the difference is for a person working in the United States.
Your company is currently a small company but is expanding rapidly.


```{r}
require("tidyverse")
require("dplyr")
ds_salaries <- read.csv("C:/Users/crisf/OneDrive/Documents/DSE5002/project_1/r project data.csv")
view(ds_salaries)
```

I plan on looking up salaries by experience, employment type, company size, and remote ratio to see if/how these factors influence the salary ranges.
I'm going to want to compare what the avg US-based salaries look like vs non-US based (the employee residence)
I should see if I can determine any trends over time (perhaps look by year). Get ready for so, so many ggplot graphs.

```{r}
# First, I want to look at the salaries in USD by experience
# Just so I remember: EN Entry-level / Junior MI Mid-level / Intermediate SE Senior-level / Expert EX Executive-level

ds_salaries <- ds_salaries %>%
 mutate(experience_level = factor(recode(experience_level, 
                                  EN = "Entry Level",     # I wanted to rename these codes so they made sense to me
                                  MI = "Mid-Level",       # I also looked up how to make sure they showed up in the right order
                                  SE = "Senior Level", 
                                  EX = "Executive Level"),
                              levels = c("Entry Level", "Mid-Level", "Senior Level", "Executive Level")))

salary_by_experience <- ds_salaries %>%
  group_by(experience_level) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_experience)
```
```{r}
ggplot(salary_by_experience, aes(x = experience_level, y = avg_salary, fill = experience_level)) +
  geom_col() +
  labs(title = "Salary Distribution by Experience Level",
       x = "Experience Level",
       y = "Salary (USD)")
```
```{r}
# Next, I want to see whether employment type influences salaries
# PT Part-time / FT Full-time / CT Contract / FL Freelance

salary_by_employment <- ds_salaries %>%
  group_by(employment_type) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_employment)
```
```{r}
ggplot(salary_by_employment, aes(x = employment_type, y = avg_salary, fill = employment_type)) +
  geom_col() +
  labs(title = "Salary Distribution by Employment Type",
       x = "Employment Type",
       y = "Salary (USD)")
```
```{r}
# How might the size of the company influence salaries?

salary_by_company_size <- ds_salaries %>%
  group_by(company_size) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_company_size)
```
```{r}
ggplot(salary_by_company_size, aes(x = company_size, y = avg_salary, fill = company_size)) +
  geom_col() +
  labs(title = "Average Salary by Company Size",
       x = "Company Size",
       y = "Salary (USD)")
```

```{r}
# What about whether the person works in the office, remotely, or hybrid?

salary_by_remote_ratio <- ds_salaries %>%
  group_by(remote_ratio) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_remote_ratio)

```
```{r}
ggplot(salary_by_remote_ratio, aes(x = remote_ratio, y = avg_salary, fill = remote_ratio)) +
  geom_col() +
  labs(title = "Average Salary by Remote Ratio",
       x = "Remote Ratio",
       y = "Salary (USD)")

ggplot(salary_by_remote_ratio, aes(x = remote_ratio, y = avg_salary, fill = remote_ratio)) +
  geom_col() +
  geom_point(aes(y = min_salary), color = "yellow") +  
  geom_point(aes(y = max_salary), color = "green") +  
  geom_point(aes(y = median_salary), color = "orange") 
  labs(title = "Average Salary by Remote Ratio with Min, Max, and Median",
       x = "Remote Ratio (%)",
       y = "Salary (USD)")


# That's not how I thought the salaries would trend!
```
```{r}
# Does employee residence factor into salaries?
salary_by_residence <- ds_salaries %>%
  group_by(employee_residence) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_residence)
```

```{r}
ggplot(salary_by_residence, aes(x = employee_residence, y = avg_salary, fill = employee_residence)) +
  geom_col() +
  labs(title = "Average Salary by Employee Residence",
       x = "Employee Residence",
       y = "Salary (USD)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

#This is an overwhelming plot. I'm going to try something else.

```

```{r}
# Can I see the top 50 highest-paid people and where they happen to reside, instead?

top_50_salaries <- ds_salaries %>%
  arrange(desc(salary_in_usd)) %>%
  head(50) 

ggplot(top_50_salaries, aes(x = reorder(employee_residence, salary_in_usd), y = salary_in_usd, fill = employee_residence)) +
  geom_bar(stat = "identity") +
  labs(title = "Highest Salaries by Employee Residence",
       x = "Employee Residence",
       y = "Salary (USD)")

#I don't think I gleaned anything useful from this, but I'll keep it in here, just the same.

```
```{r}
# Looking at just the countries with the highest average salary

top_10_salaries <- salary_by_residence %>%
  arrange(desc(avg_salary)) %>%
  head(10)

ggplot(top_10_salaries, aes(x = reorder(employee_residence, avg_salary), y = avg_salary, fill = employee_residence)) +
  geom_col() +
  labs(title = "Top 10 Countries with the Highest Average Salary",
       x = "Employee Residence",
       y = "Salary (USD)")
```
```{r}
# Now I'm just being nosy

btm_10_salaries <- salary_by_residence %>%
  arrange(avg_salary) %>%
  head(10)

ggplot(btm_10_salaries, aes(x = reorder(employee_residence, avg_salary), y = avg_salary, fill = employee_residence)) +
  geom_col() +
  labs(title = "Countries with the Lowest Average Salary",
       x = "Employee Residence",
       y = "Salary (USD)")
```


```{r}
# I want to see how US salaries compared with non-US salaries overall

us_data <- ds_salaries %>% filter(employee_residence == "US")
offshore_data <- ds_salaries %>% filter(employee_residence != "US")

us_salary_avg <- mean(us_data$salary_in_usd)
offshore_salary_avg <- mean(offshore_data$salary_in_usd)

print(us_salary_avg)
print(offshore_salary_avg)
```

```{r}
salary_comparison <- data.frame(      # I asked the internet for help fixing this one up
  Region = c("US", "Offshore"),
  Avg_Salary = c(us_salary_avg, offshore_salary_avg)
)

ggplot(salary_comparison, aes(x = Region, y = Avg_Salary, fill = Region)) +
  geom_bar(stat = "identity") +
  labs(title = "Salary Comparison: US vs. Offshore",
       x = "Region",
       y = "Average Salary (USD)")
```

```{r}
# Does job title make a difference?

salary_by_job_title <- ds_salaries %>%
  group_by(job_title) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_job_title)
```

```{r}
ggplot(salary_by_job_title, aes(x = job_title, y = avg_salary, fill = job_title)) +
  geom_col() +
  labs(title = "Average Salary by Job Title",
       x = "Job Title",
       y = "Salary (USD)")

#I can't really tell by this graph
#Also didn't realize there were so many job titles
#I think cleaning this data is a little beyond my scope at the moment
```
```{r}
# I asked ChatGPT for alternative ways to display this data

library(knitr)
kable(salary_by_job_title, caption = "Average Salary by Job Title")
```

```{r}
# I want to look at salary trends over time, now that I have a basic idea of how various factors have influenced average salaries
# This chart calculates the average, median, minimum, and maximum salaries for all individuals in the data frame by the year each salary was paid

salary_by_year <- ds_salaries %>%
  group_by(work_year) %>%
  summarize(
    avg_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    min_salary = min(salary_in_usd),
    max_salary = max(salary_in_usd)
  )

print(salary_by_year)

```

```{r}
ggplot(salary_by_year, aes(x = work_year, y = avg_salary)) +
  geom_line() +
  labs(title = "Salary Trends Over Time",
       x = "Year",
       y = "Average Salary (USD)")

# and that is a massive leap from 2021 to 2022!
```
```{r}
# I wanted to create a plot that also showed the minimum, maximum, and median salaries, so I consulted about this one
# The change in average salary doesn't look as severe with this view because of the inclusion of the max salaries
# There was one individual who make $600k in 2021
# It is interesting to see that, while the average salaries are highest in 2022, the maximum salaries are the lowest of the 3 years 

ggplot(salary_by_year, aes(x = work_year)) +
  geom_line(aes(y = avg_salary, color = "Average Salary"), size = 1) +   
  geom_point(aes(y = min_salary, color = "Min Salary")) + 
  geom_point(aes(y = max_salary, color = "Max Salary")) +  
  geom_point(aes(y = median_salary, color = "Median Salary")) + 
  labs(title = "Salary Trends Over Time",
       x = "Year",
       y = "Salary (USD)",
       color = "Salary Type") +  
  scale_color_manual(values = c("Average Salary" = "blue", 
                                "Min Salary" = "orange", 
                                "Max Salary" = "green", 
                                "Median Salary" = "purple"))
```